---
title: "DeadWood_DWDDecay"
author: ""
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DeadWood_DWDDecay}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

`DeadWood_DWDDecay` manages the downed woody debris (DWD) pool. Each timestep
it receives fallen snags from `DeadWood_snagDecay`, remaps their decay class
to the DWD scale, advances all DWD records through DC1–DC5 via a Markov
transition matrix, and stochastically removes fully decomposed material.

## Inputs

| Object | Class | Description |
|--------|-------|-------------|
| `fallenSnags` | `data.table` | Snags that fell this timestep from `DeadWood_snagDecay` |

## Outputs

| Object | Class | Description |
|--------|-------|-------------|
| `DWDTable` | `data.table` | Current DWD inventory: `pixelID`, `species`, `DC`, `ageInDC`, `initBiomass` |

## Parameters

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `DWDTransMat` | `matrix` | 5×5 zero matrix | Annual DC transition probability matrix for DWD |
| `snagToDWD_DCmap` | `integer` | `1:5` | Maps snag DC at fall to starting DWD DC (length 5) |
| `DWD_lossProb` | `numeric` | `rep(0.05, 5)` | Annual complete-loss probability by DC (length 5) |

## Example

```{r example}
library(SpaDES.core)
library(data.table)

mySim <- simInit(
  times   = list(start = 0, end = 10),
  modules = list("DeadWood_DWDDecay"),
  params  = list(DeadWood_DWDDecay = list(
    DWDTransMat     = myDWDMat,   # supply a real 5x5 matrix
    snagToDWD_DCmap = c(1L, 2L, 2L, 3L, 4L),
    DWD_lossProb    = c(0.00, 0.05, 0.06, 0.14, 0.28)
  )),
  objects = list(fallenSnags = emptySnagTable),
  paths   = list(modulePath = "path/to/modules")
)

mySim <- spades(mySim)
```
