---
title: "DeadWood_DWDDecay"
author: ""
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# DeadWood_DWDDecay

A [SpaDES](https://spades.predictiveecology.org/) module that manages the
downed woody debris (DWD) pool. It receives fallen snags from
`DeadWood_snagDecay`, maps their decay class onto the DWD scale (DC1–DC4),
and advances each piece through decomposition using a time- and
diameter-dependent logistic transition model. Pieces that pass through DC4 are
removed as fully decomposed.

Default parameters are calibrated for *Pinus strobus* (Eastern White Pine) using
Vanderwel et al. (2006).

---

## Inputs

| Object | Class | Description |
|--------|-------|-------------|
| `fallenSnags` | `data.table` | Snags that fell in the current 5-year timestep, produced by `DeadWood_snagDecay`. Required columns: `pixelID` (integer), `species` (character), `DC` (integer, snag DC 1–5), `ageInDC` (integer), `initBiomass` (numeric, Mg ha⁻¹), `diameter_cm` (numeric, cm). |

**Minimum diameter:** Any piece with `diameter_cm < 7.5 cm` will cause an error.

---

## Outputs

| Object | Class | Description |
|--------|-------|-------------|
| `DWDTable` | `data.table` | Current DWD inventory, updated every 5 years. Fully decomposed pieces (exited DC4) are removed. |

`DWDTable` schema:

| Column | Type | Description |
|--------|------|-------------|
| `pixelID` | integer | Raster pixel of origin |
| `species` | character | Tree species |
| `DC` | integer | Current DWD decay class (1–4) |
| `ageInDC` | integer | Years spent in the current DWD decay class |
| `ageSinceEntry` | integer | Total years since the piece entered the DWD pool |
| `initBiomass` | numeric | Biomass at time of death (Mg ha⁻¹); never modified after entry |
| `diameter_cm` | numeric | Piece diameter (cm); never modified after entry |

**Note on the DWD decay class scale:** The DWD pool uses a 4-class scale.
DWD DC4 combines what would be snag DC4 and DC5 on the snag scale — both
represent highly decomposed material on the forest floor.

---

## Parameters

| Parameter | Type | Default | Range | Description |
|-----------|------|---------|-------|-------------|
| `DWD_logisticParams` | `data.table` | See below | — | 4-row table of logistic transition parameters. Row *i* governs the transition from DWD DC*i* to DC*i*+1 (or out of the pool for row 4). Source: paper appendix. |
| `snagToDWD_DCmat` | matrix | See below | — | 5×4 probability matrix mapping snag DC at fall (rows 1–5) to DWD DC entered (columns 1–4). Source: Vanderwel et al. (2006) Table 4. |

### Default logistic parameters (`DWD_logisticParams`)

```{r logisticParams}
data.table::data.table(
  a0 = c(-7.295, -9.663,  3.882, -5.584),
  a1 = c( 1.888,  0.000, -2.986,  0.000),
  b0 = c( 1.325,  0.934, -0.076,  0.149),
  b1 = c(-0.239,  0.000,  0.093,  0.000)
)
# Row 1: DC1 → DC2;  Row 2: DC2 → DC3
# Row 3: DC3 → DC4;  Row 4: DC4 → out (fully decomposed)
```

Parameters `a` and `b` in the logistic model are piece-specific and depend on
diameter (see formula below).

### Default snag-to-DWD decay class mapping (`snagToDWD_DCmat`)

Rows are snag DC at time of fall; columns are DWD DC entered:

```{r snagToDWD}
matrix(c(
  0.000, 0.162, 0.815, 0.023,  # snag DC1
  0.000, 0.078, 0.871, 0.052,  # snag DC2
  0.000, 0.036, 0.854, 0.111,  # snag DC3
  0.000, 0.016, 0.762, 0.222,  # snag DC4
  0.000, 0.007, 0.599, 0.395   # snag DC5
), nrow = 5, byrow = TRUE,
dimnames = list(paste0("snagDC", 1:5), paste0("DWDDC", 1:4)))
```

Regardless of snag DC, fallen snags nearly always enter the DWD pool at DC2 or
DC3. DC1 entry probability is zero for all snag classes, reflecting the
observation that snags are already partially decomposed by the time they fall.

---

## Events

The module fires three event types.

### `init` (once, at simulation start)

- Validates that `DWD_logisticParams` is a 4-row `data.table` and
  `snagToDWD_DCmat` is a 5×4 matrix.
- Creates an empty `DWDTable` with the correct schema.
- Schedules `receive` (priority 2) and `transition` (priority 3) at
  `start(sim) + 5`.

### `receive` (every 5 years, priority 2)

Runs before `transition` within the same timestep. Accepts any fallen snags
from `sim$fallenSnags`:

1. For each piece, samples a DWD entry DC from the row of `snagToDWD_DCmat`
   corresponding to the piece's snag DC:

$$\text{DWD-DC}_\text{entry} \sim \text{Categorical}\!\left(\text{snagToDWD\_DCmat}[\text{snag-DC},\, \cdot\,]\right)$$

2. Sets `ageInDC = 0` and `ageSinceEntry = 0` for all incoming pieces.
3. Appends the new pieces to `DWDTable`.

If `fallenSnags` is empty or `NULL`, `receive` does nothing.

### `transition` (every 5 years, priority 3)

Advances all pieces currently in `DWDTable` through the logistic decay model:

**Step 1 — Compute piece-specific logistic parameters**

For each piece, parameters `a` and `b` are derived from piece diameter using the
logistic parameter row for the piece's current DC:

$$a = a_0 + a_1 \cdot \ln(D)$$
$$b = b_0 + b_1 \cdot \ln(D)$$

where $D$ is `diameter_cm` and $(a_0, a_1, b_0, b_1)$ come from the row of
`DWD_logisticParams` indexed by the piece's current DC.

**Step 2 — Logistic survival function**

The cumulative probability that a piece has *not yet* transitioned to the next
DC by age $t$ (years since DWD entry) is given by:

$$S(t) = \frac{1}{1 + e^{a + b \cdot t}}$$

**Step 3 — Conditional transition probability**

For a piece at age $A$ (= `ageSinceEntry`), the probability that it transitions
during the next 5-year interval $[A,\, A+5]$, given that it has not yet
transitioned, is:

$$p_\text{transition} = \frac{S(A) - S(A+5)}{S(A)}$$

This is the conditional probability: the fraction of pieces that survive to age
$A$ but not to age $A+5$. The result is clamped to $[0, 1]$. If $S(A) \approx 0$
(piece is essentially certain to have transitioned already), the probability is
set to 1.

**Step 4 — Stochastic transition and removal**

Each piece independently draws a transition outcome:

$$\text{Transition} \sim \text{Bernoulli}(p_\text{transition})$$

- Pieces that transition: `DC` increments by 1; `ageInDC` resets to 0.
- Pieces that do not transition: `DC` unchanged; `ageInDC` increments by 5.
- All pieces: `ageSinceEntry` increments by 5 regardless.
- Pieces with `DC > 4` are removed from `DWDTable` (fully decomposed).

---

## Event scheduling and module interactions

`DeadWood_Mortality` fires every year; the remaining modules fire every 5 years.
Priority determines order only among events scheduled at the same simulation time.

| Timestep | Priority | Module | Event | Purpose |
|----------|----------|--------|-------|---------|
| every 1 yr | 0 | `DeadWood_Mortality` | `transition` | Apply baseline mortality, append to `cohortData` |
| every 5 yr | 1 | `DeadWood_snagDecay` | `transition` | Advance snag DC, produce `fallenSnags` |
| every 5 yr | 2 | `DeadWood_DWDDecay` | `receive` | Accept `fallenSnags` into DWD pool ← **this module** |
| every 5 yr | 3 | `DeadWood_DWDDecay` | `transition` | Advance DWD DC via logistic model ← **this module** |
| every 5 yr | 4 | `DeadWood_Biomass` | `transition` | Compute biomass rasters from updated pools |
| every 5 yr | 5 | `DeadWood_Biomass` | `plot` | Capture biomass snapshots |

The `receive` event (priority 2) fires before `transition` (priority 3) within
the same timestep, ensuring newly fallen snags are incorporated before any
DC advancement occurs.

---

## Usage example

```{r usage}
library(SpaDES.core)

# fallenSnags would normally be supplied by DeadWood_snagDecay
fallenSnags <- data.table::data.table(
  pixelID     = c(1L, 1L),
  species     = "Pinus strobus",
  DC          = c(2L, 3L),        # snag DC at time of fall
  ageInDC     = c(10L, 5L),
  initBiomass = c(12.5, 8.3),     # Mg ha-1
  diameter_cm = c(22.0, 18.5)
)

mySim <- simInit(
  times   = list(start = 0, end = 50),
  params  = list(),
  modules = list("DeadWood_DWDDecay"),
  objects = list(fallenSnags = fallenSnags)
)

mySim <- spades(mySim)
mySim$DWDTable
```

---

## References

Vanderwel, M.C., Malcolm, J.R., Smith, S.M., and Islam, N. (2006).
An integrated model for snag and downed woody
debris decay class transition.
*Forest Ecology and Management*, 234(1–3), 48–59.
<https://doi.org/10.1016/j.foreco.2006.06.020>

---

## Package dependencies

- [`SpaDES.core`](https://github.com/PredictiveEcology/SpaDES.core) (>= 3.0.0)
- [`data.table`](https://CRAN.R-project.org/package=data.table)
